---
title: "RiboSeeker Guide"
author: "Ning Zhang"
date: "7/21/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE)
```

<hr>

### Introduction

RiboSeeker is an R package for ribosome profiling data analysis. It has two components:

* [Ribomake](https://github.com/nzhang89/Ribomake): a Snakemake workflow to process ribosome 
profiling data from raw reads
* [RiboSeeker](https://github.com/nzhang89/RiboSeeker): an R package for further data analysis
from aligned reads

<br>

### Quick Start

Here, we focus on using the RiboSeeker R package for Ribo-seq data analysis. The example data
used are from a human K562 ribosome profiling dataset
([https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM3566410](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM3566410)).
We run through the Ribomake workflow using hg38 human genome and Ensembl 98 annotation.

<br>

#### Set up packages and data

First install and load RiboSeeker and several other packages to run the example data.

```{r package}
library(GenomicAlignments) # deals with bam alignments
library(GenomicFeatures) # deals with genome annotations
library(GenomicRanges) # deals with general genomic data
library(ggplot2) # deals with visualizations

library(RiboSeeker)
```

Then specify the example data files. We only take the reads aligned to chromosome 1 to minimize
the running time and storage space. For genome annotation, we also only take chromosome 1
annotations.

```{r data}
# example data folder
example_folder = system.file('extdata', package='RiboSeeker')

# bam file
bam_file = file.path(example_folder, 'GSM3566410.chr1.sorted.bam')
# genome annotation txdb file
txdb_file = file.path(example_folder, 'hg38.Ens_98.chr1.txdb.sqlite')

# # alternatively, you can use bioconductor human annotation package
# if (!requireNamespace('BiocManager', quietly=TRUE)) {
#   install.packages("BiocManager")
# }
# BiocManager::install('TxDb.Hsapiens.UCSC.hg38.knownGene')
# library(TxDb.Hsapiens.UCSC.hg38.knownGene)
```

Next load BAM and TxDb files. We use `loadBam` function to load an indexed bam file and return
a `GAlignments` object of aligned reads. The `loadBam` function is essentially a wrapper of
`readGAlignments` function in `GenomicAlignments` package. **Note** that Ribo-seq data are 
typically single-end, the paired RNA-seq data can be paired-end. Since the single- or
paired-end information is not very important in Ribo-seq data analysis, we simply treat all
BAM files as single-end. A message will be printed out if paired-end BAM file is detected.

```{r load_data}
# load bam file
bam = loadBam(bam_file)
bam

# load txdb file
txdb = loadDb(txdb_file)

# alternatively, you can use bioconductor human annotation txdb object
# txdb = TxDb.Hsapiens.UCSC.hg38.knownGene
```

<br>

#### Read length distribution

Ribo-seq reads, or ribosome protected fragments (RPFs) tend to have a specific length 
distribution. Usually the RPFs are around 30 bp. We implement `readLengthDist` function to 
count the number and percentage of reads for each read length. The function 
`readLengthDistPlot` can be used to plot the read length distribution (percentage or actual
read counts) as a barplot.

```{r read_len_dist}
# read length distribution
read_len_dist = readLengthDist(bam)
read_len_dist
```

```{r read_len_dist_plot, fig.width=7, fig.height=4}
# read length distribution plot
p = readLengthDistPlot(read_len_dist, color='steelblue', type='pct')
p
```

If you run through the Ribomake workflow, it will also generate a read length distribution
plot. You might notice the distribution is somehow different than presented here. There might
be two reasons: 1. the example data presented here only include reads aligned to chromosome 1;
and 2. here we take all aligned reads regardless of where they are mapped (for example, reads
aligned to intergenic regions are also counted here). But the Ribomake workflow only takes 
reads that are aligned to transcripts.

<br>

#### Genomic feature distribution of aligned reads

It may also be interesting to see where the reads are aligned on the genome. We implement 
`readGenomeDist` function to assign reads to genomic features such as CDS, UTR, Intron, and
Intergenic. This function resizes reads to its 5'-end positions to minimize overlap with 
multiple features. It returns a list of two elements, the first is the genomic feature
assignment for each read, and the second is the summary statistics for user specified genomic
features.

```{r read_feature_dist}
# read genomic feature distribution
read_feature_dist = readGenomeDist(bam, txdb, category=c('CDS', 'UTR', 'Intron', 'Intergenic'), 
  fiveEndOnly=TRUE, ignoreStrand=TRUE)
read_feature_dist$summary
```

```{r read_feature_dist_plot, fig.width=7, fig.height=2}
# read genomic feature distribution plot
p = readGenomeDistPlot(read_feature_dist$summary, type='pct')
p
```

We can see that for this example dataset, most reads are assigned to CDS and UTR.
What is interesting about the full dataset is that if we take all the reads aligned to all
chromosomes, the most majority of reads are actually assigned to intergenic regions.

<br>

#### Metagene

A very useful and important plot in Ribo-seq data analysis is the metagene plot, especially
around CDS start (start codon) and CDS end (stop codon) regions. Due to the large size of the
ribosome, the 5'-end of RPFs are actually upstream of the ribosome P-site. This shift is called
**P-site offset**. Metagene plot is a good way to determine the P-site offset.

We implement `calcMetagene` and `metagenePlot` function to compute the metagene matrix and make
the plot. For metagene matrix calculation, we require a BAM alignment and a metagene region.
Similar with genomic feature plot, we also take only the 5'-end position for each read. If the
region is not specified, we will extract all CDS start and end regions from the TxDb
annotation. Then the metagene matrix is calculated using either user specified region or the 
extracted two CDS start and end regions. Each row in the metagene matrix represents a
transcript, and each column represents a position in the metagene region. We also provide the
option to normalize row-wise the counts to the total counts in each row.

For metagene plot, one plot will be made if the metagene region is specified by users. 
Otherwise, two plots around CDS start and end regions will be made. Position 0 in CDS start
means the first base of the start codon, and position 0 in CDS end means the last base of the
stop codon.

```{r metagene}
# metagene
metagene = calcMetagene(bam, txdb=txdb)
```

```{r metagene_plot, fig.width=13, fig.height=5}
# metagene plot
p = metagenePlot(metagene, metageneFun='mean')
p
```

From the metagene plot, we do not specify the metagene region and use the extracted CDS start
and end regions. We can see a large peak at -12 position in CDS start regions. Therefore, we
can determine that the P-site offset for this example dataset is 12 bp. In fact, the 12 bp 
offset is a very common number for many Ribo-seq experiments. For CDS end regions, we can also
see a large peak at -17 position. This also makes sense, because if we shift the reads 12 bp
downstream, this peak will be at -5 position, which is exactly the P-site position since 
position -5, -4, -3 is the last codon before stop codon, and position -2, -1, and 0 is the
stop codon.

<br>

#### Shift reads

From the read length distribution and the metagene plots, we can determine the most abundant
read lengths and the P-site offset. For this example dataset, we can take 28 bp reads with 12
bp P-site offset. Before downstream analysis, reads needs to be shifted. We implement 
`shiftReads` function to select certain read lengths and shift reads towards downstream
direction.

```{r shift_read}
# shift reads
# here we do not select read length but shift all reads 12 bp downstream
# set fiveEndOnly to FALSE will return a GAlignments object instead of a GRanges object
bam_shift = shiftReads(bam, shiftLens=12, fiveEndOnly=FALSE)
bam_shift
```

We can also make a metagene plot using the shifted reads. In this case, we can see the large
peak should be aligned at exactly position 0, which is the start codon.

```{r shift_read_metagene_plot, fig.width=13, fig.height=5}
# metagene plot
p = metagenePlot(calcMetagene(bam_shift, txdb=txdb), metageneFun='mean')
p
```

<br>

#### Further analysis

<br>

## Session Info
```{r session, echo=FALSE, message=FALSE, warning=FALSE}
sessionInfo()
```
